Evaluating Perceptual Judgements on 3D Printed Bar Charts

### 2024-01-22 jds.rmd
Updates and Reviewer 1 corrections

---
title: "Evaluating Perceptual Judgements on 3D Printed Bar Charts"
author:
- first: Tyler
  first-init: T.
  last: Wiederich
  marker: 1
  footnote-id: 1
  email: twiederich2@huskers.unl.edu
- first: Susan
  first-init: S.
  last: VanderPlas
  marker: 1
  email: susan.vanderplas@unl.edu
short-title: Perceptual Judgements on 3D Printed Bar Charts
output:
  jds.rmd::pdf_article: 
    extra_dependencies: "subfig"
affiliation:
- institution: "University of Nebraska-Lincoln"
  prefix: Department of Statistics
  country: United States of America
  marker: 1
classoption:
- letterpaper
- inpress
- dvipsnames
preamble: |
  \usepackage{amsfonts,amsmath,amssymb,amsthm} 
  \usepackage{booktabs} 
  \usepackage{lipsum} 
  \PassOptionsToPackage{dvipsnames}{xcolor} 
  \newcommand{\svp}[1]{{\textcolor{ForestGreen}{#1}}} 
  \newcommand{\tw}[1]{{\textcolor{Red}{#1}}}
abstract: "Graphical design principles typically recommend minimizing the dimensionality of a visualization - for instance, using only 2 dimensions for bar charts rather than providing a 3D rendering, because this extra complexity may result in a decrease in accuracy. This advice has been oft repeated, but the underlying experimental evidence is focused on fixed 2D projections of 3D charts. In this paper, we describe an experiment which attempts to establish whether the decrease in accuracy extends to 3D virtual renderings and 3D printed charts. We replicate the grouped bar chart comparisons in the  1984 Cleveland & McGill study, assessing the accuracy of numerical estimates using different types of 3D and 2D renderings."
keywords:
- graphics
- 3D bar charts
- 3D printing
bibliography: JDS2023.bib
footnote:
- content: Tyler Wiederich
  type: corresp
  id: 1
---

```{r, include = F}
knitr::opts_chunk$set(echo = F, dpi = 300)

@@

res = results %>% 
  left_join(trueRatios, by = 'fileID') %>% 
  mutate(response = log2(abs(byHowMuch - Height*100) + 1/8),
         subject = paste0(nickname, participantUnique),
         ratioLabel = round(100*Height, 1)) %>% 
  filter(subject %in% valid.users$subject) %>% 
  arrange(appStartTime) %>% 
  filter(whichIsSmaller == 'Triangle (▲)',
         as.Date.POSIXct(appStartTime) <= '2023-05-22') #Only verified results from before the start of SDSS

@@

The first experiment in @clevelandGraphical1984 (the position-length experiment), used five types of bar charts: two types of grouped bar charts and three types of stacked bar charts.
Each chart had two bars marked for comparison; participants were asked to determine which bar was smaller and give the  perceived ratio of the smaller bar to the larger bar. @fig:bar-types show the two types of grouped bar charts. 
We are primarily interested in the grouped bar charts (in part because 3D printing is not yet inexpensive enough to make moderate-scale stacked bar chart experiments viable), which consisted of two comparison bars which were either adjacent or in separate groups.
These grouped bar charts will be referenced as adjacent and separated graph types in this paper, respectively.

```{r bar-types}
#| message: false
#| error: false
#| warning: false
#| fig-width: 4
#| fig-height: 4
#| fig-cap: "@clevelandGraphical1984 used two different types of grouped bar charts, which are replicated here."
#| fig-subcap: 
#|    - "Adjacent"
#|    - "Separated"
#| layout-ncol: 2
library(ggplot2)

datasets$data[[1]] %>%
  ggplot(aes(x = Group, y = Height, group = GroupOrder)) + 
  geom_col(position = "dodge", fill = "white", color = "black") + 
  geom_text(aes(x = Group, y = 5, group = GroupOrder, label = IDchr), position = position_dodge(0.9)) + 
  theme_bw() + 
  theme(axis.title = element_blank(), axis.text.x = element_blank()) + 
  ggtitle("Adjacent Chart")

datasets$data[[1]] %>%
  ggplot(aes(x = factor(Order %%2), y = Height, group = GroupOrder)) + 
  geom_col(position = "dodge", fill = "white", color = "black") + 
  geom_text(aes(x = factor(Order %%2), y = 5, group = GroupOrder, label = IDchr), position = position_dodge(0.9)) + 
  theme_bw() + 
  theme(axis.title = element_blank(), axis.text.x = element_blank()) + 
  ggtitle("Separated Chart")

@@

```{r sine-illusion}
#| message: false
#| error: false
#| warning: false
#| cache: false
#| fig-width: 4
#| fig-height: 4
#| out-width: "30%"
#| fig-cap: "An illustration of the sine illusion\\citep{vanderplasSignsSineIllusion2015}, also known as the line-width illusion. All vertical lines are the same length, but the lines in the middle of the curve appear to be much shorter. The illusion results when implicit perceptual corrections useful for perceiving the size of objects with depth are applied to 2D stimuli with no actual depth. As 3D heuristics occasionally cause inaccurate communication when applied to 2D objects, it is reasonable to think that there might be some situations where charts making use of a realistic third dimension might be perceived more accurately than their 2D equivalents."
#| fig-subcap:
#|    - "Sine Illusion"
#|    - "3D surface with similar features"
#|    - "Perceived vs. actual length"
#' Function to data frame
#'  
#' @param n number of values 
#' @param ell extent of vertical range
#' @param x range of horizontal values
#' @param f function
#' @param fp first derivative of function f
#' @param f2p second derivative of function f
#' @return data frame with n function values and derivatives along the x axis for a range given by x
#' @example
#' f <- function(x) 2*sin(x)
#' fp <- function(x) 2*cos(x)
#' f2p <- function(x) -2*sin(x)
#' dframe <- function.frame(n=50, ell=1, x=c(0,2*pi), f, fp, f2p)
#' require(ggplot2)
#' qplot(x, y=ystart, yend=yend, xend=x, geom="segment", data=dframe) +
#'    geom_point(aes(x,y))
function.frame <- function(n=200, ell=1, x, f, fprime, f2prime) {
  x <- seq(x[1], x[2], length=n)
  if (length(ell) != length(x)) ell <- rep(ell, length(x))
  y <- f(x)
  ystart <- y - ell/2
  yend <- y + ell/2
  # now correct for line illusion in vertical direction
  dy <- diff(range(y))
  dyl <- diff(range(c(ystart, yend)))
  # fprime and f2prime are sensitive to the aspect ratio of a plot
  # we represent it in framework of dy and dy+len
  # needs to be fixed by factor a:  
  a <- dy/dyl

@@

persp(x, z, data.persp,  xlab="", ylab="", zlab="", theta=0, phi=45, border="black", shade=.35, col="white", xlim=c(-pi/12, 2*pi+pi/12), ylim=c(-1.75, 1.75), scale=FALSE, box=FALSE, expand=3/pi, d=50) # , ltheta=0, lphi=-15

@@

There is thus good reason to be wary of the use of three dimensions where only two are necessary to convey data. 
However, the situation is different now than it was in 1984 when Cleveland \& McGill published their seminal work; it has even changed since Heer's replication study [-@heerCrowdsourcingGraphicalPerception2010b] in 2010.  
Digital graphics has developed quickly, along with the hardware necessary to support these software developments. 
As a result, we have much more natural rendering of 3D objects virtually, and we can also print graphics in three dimensions, moving artificial charts into a more natural, physical setting.
As a result, it is reasonable to reconsider the use of 3D charts, not only because of new technological developments, but also because these charts provide the opportunity to make data graphics accessible to those with limited or absent vision [@fleetHowTactileGraphics2023].

@@

Each graph is presented so that there are ten bars where only two of the bars are marked for identification, \svp{as shown in @fig:bar-types.}
Cleveland and McGill did not specify the random process for the heights of the eight other bars, so we used a scaled Beta distribution with parameters that limit excessive noise around the bars used for comparisons.
Code to reproduce the data generation process, data underlying the plots used in this study, the rendered plots and STL files, anonymized user data, and analysis code can found be at [https://github.com/TWiedRW/2023-JDS-3dcharts](https://github.com/TWiedRW/2023-JDS-3dcharts).

@@

The graphs share a common layout across all formats, where two groupings of five bars are identified by "A" and "B", respectively, and circles and triangles are used to identify the bars participants should compare. Example graphs are shown in
Figure \@ref(fig:plotTypes).
There are some graphical elements that cannot be easily portrayed via 3D printing. For this reason, all graph types do not have axes, grid lines, or floating titles.
<!-- The aspect ratio of the plots presented in the original paper is approximately 0.825:1. The aspect ratio of the plots presented in our study is closer to 0.864:1, but -->
<!-- The aspect ratio of the 2d plots (including only the bars is 2:1), while the aspect ratio of the 3d printed charts is 11:9.5. This is something we probably need to fix.... Until then, let's not go there -->

```{r plotTypes, fig.cap = "Two dimensional, two-dimensional digital rendering, and 3D-printed charts used in this study.", out.width = "100%"}
knitr::include_graphics("_images/plot-types.png", dpi = 300)
```

@@

Participants were provided a kit with five 3D-printed charts corresponding to five of the seven unique ratio comparisons. In the Shiny application, we asked participants for the kit ID, which was used to look up the corresponding digital renderings. This ensured that the experimental design was balanced across chart type and randomized with respect to the type of comparison (adjacent or separated).
We printed 21 kits containing five 3d-printed charts each, where 2 of the 7 ratios were randomly removed from the total set.
<!-- There are 21 kits in total so that each combination of ratios is accounted for. -->
<!-- Each graph is randomly assigned either an adjacent or separated comparison. -->

```{r studyDesign, fig.cap = "A graphical representation of the study design. Only five of the seven ratios were used in each kit where the ratios are randomly selected. A total of 21 kits were created to include all combinations of the five ratios."}
knitr::include_graphics("_images/study-design.png")
```

@@

One interesting facet of @clevelandGraphical1984 is the participant recruitment methodology: "For each experiment the subjects fell into two categories: (1) a group of females, mostly housewives, without substantial technical experience; (2) a mixture of males and females with substantial technical training and working in technical jobs. 
Most of the subjects in the position-length experiment participated in the position-angle experiment; in all cases repeat subjects judged the position-angle graphs first." 
It would seem likely that the authors recruited individuals within their respective departments as well as their wives.
In the spirit of replicating the study, members of the UNL Statistics department and their spouses, partners, and roommates were asked to participate in our study; this replicates the spirit of the original study without the implicit assumptions that graduate students and professors are (1) largely male, (2) heterosexual, and (3) have unemployed partners.

A total of 48 participants completed the study; demographics are shown in \@ref(fig:demographics).

```{r demographics, fig.cap="Demographic breakdown of participants in the study. All subjects were recruited from faculty and students in the statistics department at University of Nebraska-Lincoln.", message=FALSE, warning=FALSE, out.width='85%'}
library(patchwork)
valid.users$userAppStartTime <- as.POSIXct(valid.users$userAppStartTime,
           origin = '1970-1-1')

age.levels = sort(unique(valid.users$age))
p1 = valid.users[valid.users[,'userAppStartTime'] <= '2023-05-22',] %>% 
  group_by(age) %>% 
  summarise(count = n()) %>% 
  ggplot(mapping = aes(x = factor(age, levels = age.levels),
                       y = count)) + 
  geom_bar(stat = 'identity') +
  geom_text(mapping = aes(y = count+1, label = count),
             size = 2) +
  labs(title = 'Age Groups',
       x = '',
       y = 'Count') +
  theme_bw() + 
  theme(plot.title = element_text(hjust=0.5),
        axis.text.x = element_text(angle = 45, hjust=1, size = 8))

p2 = valid.users[valid.users[,'userAppStartTime'] <= '2023-05-22',] %>% 
  group_by(gender) %>% 
  summarise(count = n()) %>% 
  ggplot(mapping = aes(x = gender,
                       y = count)) + 
  geom_bar(stat = 'identity') +
  geom_text(mapping = aes(y = count+1, label = count),
             size = 2) +
  labs(title = 'Gender Groups',
       x = '',
       y = 'Count') +
  theme_bw() + 
  theme(plot.title = element_text(hjust=0.5),
        axis.text.x = element_text(angle = 45, hjust=1, size = 8))

educ.levels = sort(unique(valid.users$education))[c(2,4,3,1)]
p3 = valid.users[valid.users[,'userAppStartTime'] <= '2023-05-22',] %>% 
  group_by(education) %>% 
  summarise(count = n()) %>% 
  ggplot(mapping = aes(x = factor(education, levels = educ.levels),
                       y = count)) + 
  geom_bar(stat = 'identity') +
  geom_text(mapping = aes(y = count+2, label = count),
             size = 2) +
  labs(title = 'Education Levels',
       x = '',
       y = 'Count') +
  theme_bw() + 
  theme(plot.title = element_text(hjust=0.5),
        axis.text.x = element_text(angle = 45, hjust=1, size=8))

@@

@@

A Shiny applet was used for data collection, along with the provided kit of 3D printed charts.
Participants provided informed consent through the applet, and then were asked for demographic information (age, gender, education level). 
Then, participants were shown a "practice" page which allowed them to experiment with the data collection interface and practice estimating the ratio between the bars, as shown in Figure \@ref(fig:practice)

@@

Directly before the experiment started, participants were asked to provide the kit ID, along with directions indicating that if the instructions indicated that participants should use a 3D chart for a task, the participant should select a chart from the kit, enter the ID code of that chart from the bottom of the object, and complete the requested task.
Participants were also instructed to make quick judgments for each graph and not to measure or estimate ratios using physical objects.

@@

Each graph (or prompt, in the case of 3D printed charts) in the applet had two corresponding questions for participants to answer: first, participants were to identify the smaller bar by shape, and then, participants were to estimate the ratio of the size of the smaller bar to the size of the larger bar, as shown in Figure \@ref(fig:experiment3dRender).

@@

All responses that incorrectly identified the smaller bar were removed from the study before analysis.

## Midmeans of Log Absolute Errors
Cleveland and McGill used 
$$\log_2(|\text{Judged Percent} - \text{True Percent}|+1/8)$$
to measure accuracy of their participant's responses. 
In their study, log base 2 seemed appropriate due to "average relative errors changing by factors less than 10."
They also added 1/8 to prevent distortions when the errors were close to zero. 
@heerCrowdsourcingGraphicalPerception2010b followed the same analysis method, replicating many (but not all) of the results presented in the original paper.

Figure \@ref(fig:midmeans-log-errors) shows the midmeans of the log absolute errors compared to the true ratio of the bars for each graph type and comparison type. 
The results suggest that the log absolute errors increase for greater differences between the smaller and larger bars, but are consistent across the graph types. In addition, there is a tendency

@@

res %>% 
  mutate(type = "Raw Data") %>%
  bind_rows(tmp) %>%
  # group_by(Height, graphtype, plot) %>% 
  ggplot(mapping = aes(x = Height*100, y = response,
                       color = graphtype, fill = graphtype)) +
  geom_smooth(se = T, alpha=.125,  expand = F, method = "loess") +
  geom_point(aes(shape = type, alpha = type, size = type)) +
  scale_shape_manual("Data", values = c(19, 1)) + 
  scale_size_manual("Data", values = c(2, 1)) + 
  scale_alpha_manual("Data", values = c(1, .25)) + 
  scale_color_discrete(labels = c('Adjacent', 'Separated')) +
  scale_x_continuous(limits = c(0, 100)) + 
  facet_wrap(~plot, labeller = labeller(plot = plot.types)) +
  labs(title = '',
       x = 'True Proportional Difference (%)',
       y = 'Log Error',
       color = 'Comparison Type') +
  theme_bw() +
  theme(legend.position = 'bottom')
```
This is somewhat different than the results in @clevelandGraphical1984 and @heerCrowdsourcingGraphicalPerception2010b; in both cases the midmean log absolute errors increased until about 55% of the true proportional difference and then decreased.
It is possible that this difference is due to the fact that we did not require an explicit numerical estimate of the ratio but instead asked participants to indicate the ratio on a slider (which is essentially a number line). 
This should reduce the cognitive load required to transition from spatial comparison to numerical comparison and then to compute the proportion, but may also have impacted the results.

@@

$$y_{ijklm}=\mu+S_i+R_j+G(R)_{(k)j}+T_l+\epsilon_{ijklm}$$

@@

- $S_i\sim N(0,\sigma^2_S)$ is the effect of the $i^{th}$ subject

@@

- $G(R)_{(k)j}$ is the effect of the $k^{th}$ graph type nested in the $j^{th}$ ratio

@@

```{r, include=FALSE}
mod = lmer(response ~ (1|subject) + ratio + type + ratio:plot,
     data = res)
anova(mod)
```

>>>>

---
title: "Evaluating Perceptual Judgements on 3D Printed Bar Charts"
author:
- first: Tyler
  first-init: T.
  last: Wiederich
  marker: 1
  footnote-id: 1
  email: twiederich2@huskers.unl.edu
- first: Susan
  first-init: S.
  last: VanderPlas
  marker: 1
  email: susan.vanderplas@unl.edu
short-title: Perceptual Judgements on 3D Printed Bar Charts
output:
  jds.rmd::pdf_article: 
    extra_dependencies: "subfig"
affiliation:
- institution: "University of Nebraska-Lincoln"
  prefix: Department of Statistics
  country: United States of America
  marker: 1
classoption:
- letterpaper
- inpress
- dvipsnames
preamble: |
  \usepackage{amsfonts,amsmath,amssymb,amsthm} 
  \usepackage{threeparttable}
  \usepackage{cleveref}
  \usepackage{booktabs} 
  \usepackage{lipsum} 
  \PassOptionsToPackage{dvipsnames}{xcolor} 
  \newcommand{\svp}[1]{{\textcolor{ForestGreen}{#1}}} 
  \newcommand{\tw}[1]{{\textcolor{Red}{#1}}}
abstract: "Graphical design principles typically recommend minimizing the dimensionality of a visualization - for instance, using only 2 dimensions for bar charts rather than providing a 3D rendering, because this extra complexity may result in a decrease in accuracy. This advice has been oft repeated, but the underlying experimental evidence is focused on fixed 2D projections of 3D charts. In this paper, we describe an experiment which attempts to establish whether the decrease in accuracy extends to 3D virtual renderings and 3D printed charts. We replicate the grouped bar chart comparisons in the  1984 Cleveland & McGill study, assessing the accuracy of numerical estimates using different types of 3D and 2D renderings."
keywords:
- graphics
- 3D bar charts
- 3D printing
bibliography: JDS2023.bib
footnote:
- content: Tyler Wiederich
  type: corresp
  id: 1
---

```{r, include = F}
knitr::opts_chunk$set(echo = F, dpi = 300, message =F, warning = F, error = F, dev = "ragg_png")

@@

res_all = results %>% 
  left_join(trueRatios, by = 'fileID') %>% 
  mutate(response = log2(abs(byHowMuch - Height*100) + 1/8),
         subject = paste0(nickname, participantUnique),
         ratioLabel = round(100*Height, 1)) %>% 
  filter(subject %in% valid.users$subject) %>% 
  arrange(appStartTime)

res <- res_all %>% 
  filter(whichIsSmaller == 'Triangle (▲)',
         as.Date.POSIXct(appStartTime) <= '2023-05-22') #Only verified results from before the start of SDSS

@@

The first experiment in @clevelandGraphical1984 (the position-length experiment), used five types of bar charts: two types of grouped bar charts and three types of stacked bar charts.
Each chart had two bars marked for comparison; participants were asked to determine which bar was smaller and give the  perceived ratio of the smaller bar to the larger bar. \Cref{fig:bar-types} shows the two types of grouped bar charts. 
We are primarily interested in the grouped bar charts (in part because 3D printing is not yet inexpensive enough to make moderate-scale stacked bar chart experiments viable), which consisted of two comparison bars which were either adjacent or in separate groups.
These grouped bar charts will be referenced as adjacent and separated graph types in this paper, respectively.

```{r bar-types}
#| fig-width: 4
#| fig-height: 4
#| fig-cap: "\\citet{clevelandGraphical1984} used two different types of grouped bar charts: comparisons between adjacent bars, and comparisons between separated bars. It is widely acknowledged that comparisons between separated bars (Type 3 comparisons, in Cleveland \\& McGill's terminology) are more difficult and error-prone."
#| out-width: ".49\\linewidth"
#| fig-show: "hold"
#| layout-ncol: 2
library(ggplot2)
library(ragg)
datasets$data[[1]] %>%
  ggplot(aes(x = Group, y = Height, group = GroupOrder)) + 
  geom_col(position = "dodge", fill = "white", color = "black") + 
  geom_point(aes(x = Group, y = 5, group = GroupOrder, shape = IDchr), size = 3, position = position_dodge(0.9)) + 
  guides(shape = "none") + 
  theme_bw() + 
  theme(axis.title = element_blank(), axis.text.x = element_blank()) + 
  ggtitle("Adjacent (Type 1)")

datasets$data[[1]] %>%
  ggplot(aes(x = factor(Order %%2), y = Height, group = GroupOrder)) + 
  geom_col(position = "dodge", fill = "white", color = "black") + 
  geom_point(aes(x = factor(Order %%2), y = 5, group = GroupOrder, shape = IDchr), size = 3, position = position_dodge(0.9)) +
  guides(shape = "none") + 
  theme_bw() + 
  theme(axis.title = element_blank(), axis.text.x = element_blank()) + 
  ggtitle("Separated (Type 3)")

@@

```{r sine-illusion}
#| message: false
#| error: false
#| warning: false
#| cache: false
#| fig-width: 4
#| fig-height: 4
#| out-width: "33%"
#| fig-cap: "An illustration of the sine illusion\\citep{vanderplasSignsSineIllusion2015}, also known as the line-width illusion. All vertical lines are the same length, but the lines in the middle of the curve appear to be much shorter. The illusion results when implicit perceptual corrections useful for perceiving the size of objects with depth are applied to 2D stimuli with no actual depth. As 3D heuristics occasionally cause inaccurate communication when applied to 2D objects, it is reasonable to think that there might be some situations where charts making use of a realistic third dimension might be perceived more accurately than their 2D equivalents."
#| fig-subcap:
#|    - "Sine Illusion"
#|    - "3D surface with similar features"
#|    - "Perceived vs. actual length"
#' Function to data frame
#'  
#' @param n number of values 
#' @param ell extent of vertical range
#' @param x range of horizontal values
#' @param f function
#' @param fp first derivative of function f
#' @param f2p second derivative of function f
#' @return data frame with n function values and derivatives along the x axis for a range given by x
#' @example
#' f <- function(x) 2*sin(x)
#' fp <- function(x) 2*cos(x)
#' f2p <- function(x) -2*sin(x)
#' dframe <- function.frame(n=50, ell=1, x=c(0,2*pi), f, fp, f2p)
#' require(ggplot2)
#' qplot(x, y=ystart, yend=yend, xend=x, geom="segment", data=dframe) +
#'    geom_point(aes(x,y))
function.frame <- function(n=200, ell=1, x, f, fprime, f2prime) {
  x <- seq(x[1], x[2], length=n)
  if (length(ell) != length(x)) ell <- rep(ell, length(x))
  y <- f(x)
  ystart <- y - ell/2
  yend <- y + ell/2
  # now correct for line illusion in vertical direction
  dy <- diff(range(y))
  dyl <- diff(range(c(ystart, yend)))
  # fprime and f2prime are sensitive to the aspect ratio of a plot
  # we represent it in framework of dy and dy+len
  # needs to be fixed by factor a:  
  a <- dy/dyl

@@

par(mar = c(0, 0, 0, 0))
persp(x, z, data.persp,  xlab="", ylab="", zlab="", theta=0, phi=45, border="black", shade=.35, col="white", xlim=c(-pi/12, 2*pi+pi/12), ylim=c(-1, 1), scale=FALSE, box=FALSE, expand=2.5/pi, d=50) # , ltheta=0, lphi=-15

@@

There is thus good reason to be wary of the use of three dimensions where only two are necessary to convey data. 
However, the situation is different now than it was in 1984 when Cleveland \& McGill published their seminal work; it has even changed since Heer's replication study [-@heerCrowdsourcingGraphicalPerception2010b] in 2010.  
Digital graphics have developed quickly, along with the hardware necessary to support these software developments. 
As a result, we have much more natural rendering of 3D objects virtually, and we can also print graphics in three dimensions, moving artificial charts into a more natural, physical setting.
As a result, it is reasonable to reconsider the use of 3D charts, not only because of new technological developments, but also because these charts provide the opportunity to make data graphics accessible to those with limited or absent vision [@fleetHowTactileGraphics2023].

@@

\svp{In each graph, there are two sets of five bars each, and two of the bars are marked for identification with a circle and triangle}, \svp{as in \Cref{fig:bar-types}.}
Cleveland and McGill did not specify the random process for the heights of the eight other bars, so we used a scaled Beta distribution with parameters that limit excessive noise around the bars used for comparisons.\svp{What does excessive noise mean??}
Code to reproduce the data generation process, data underlying the plots used in this study, the rendered plots and STL files, anonymized user data, and analysis code can found be at [https://github.com/TWiedRW/2023-JDS-3dcharts](https://github.com/TWiedRW/2023-JDS-3dcharts).

@@

The graphs share a common layout across all formats, where two groupings of five bars are identified by "A" and "B", respectively, and circles and triangles are used to identify the bars participants should compare. Example graphs are shown in
\Cref{fig:plotTypes}.
There are some graphical elements that cannot be easily portrayed via 3D printing. For this reason, all graph types do not have axes, grid lines, or floating titles.
<!-- The aspect ratio of the plots presented in the original paper is approximately 0.825:1. The aspect ratio of the plots presented in our study is closer to 0.864:1, but -->
<!-- The aspect ratio of the 2d plots (including only the bars is 2:1), while the aspect ratio of the 3d printed charts is 11:9.5. This is something we probably need to fix.... Until then, let's not go there -->

```{r plotTypes, fig.cap = "Two dimensional, 3D digital rendering, and 3D-printed charts used in this study.",out.width = c("30%", "30%", "35%"), fig.show = "hold"}
knitr::include_graphics("_images/Type1-Rep01.png", dpi = 300)
knitr::include_graphics("_images/RenderedChart.png", dpi = 300)
knitr::include_graphics("_images/Kit_of_charts.png", dpi = 300)
```

@@

A diagram of the experimental design is provided in \Cref{fig:studyDesign}. The study is set up as a modified randomized incomplete block design, comprised of 3 blocks (presentation mode), 7 bar ratios (5 are selected for each participant). At each combination of ratio and presentation type, the comparison type, adjacent (Type 1) or separated (Type 3), is randomly selected. This process was used to create 21 kits of 5 3D printed charts and 10 virtual charts. There is thus a slight deviation from a completely randomized incomplete block design, in that all possible combinations of ratios, presentation modes, and comparison types are not present in our design. However, this approach provided the best experimental design given the constraints of managing physical stimuli; it took nearly a month of continuous 3D printing to create all of the charts for the kits that we assembled. A dataset containing the kit composition breakdown is available in the github repository for this project.

<!-- There are 21 kits in total so that each combination of ratios is accounted for. -->
<!-- Each graph is randomly assigned either an adjacent or separated comparison. -->

```{r studyDesign, fig.cap = "A graphical representation of the study design. Only five of the seven ratios were used in each kit; at least one of the smallest or largest ratios were included along with 4 other charts; each selected ratio was displayed in all 3 mediums. For each medium $\\times$ ratio combination, the comparison type (separated or adjacent) was randomly determined. A total of 21 kits of 3D printed charts were created to include all combinations of the five ratios."}
knitr::include_graphics("_images/design.pdf")
```

@@

One interesting facet of @clevelandGraphical1984 is the participant recruitment methodology: "For each experiment the subjects fell into two categories: (1) a group of females, mostly housewives, without substantial technical experience; (2) a mixture of males and females with substantial technical training and working in technical jobs. 
Most of the subjects in the position-length experiment participated in the position-angle experiment; in all cases repeat subjects judged the position-angle graphs first." 
It seems likely that the authors recruited individuals within their respective departments as well as their wives.
In order to replicate this atypical participant recruitment method in the modern era, members of the UNL Statistics department and their spouses, partners, and roommates were asked to participate in our study. 
This replicates the spirit of the recruitment method without the implicit assumptions that graduate students and professors are male, heterosexual, and have unemployed spouses.

A total of 48 participants completed the study; demographics are shown in \Cref{fig:demographics}.

```{r demographics, fig.cap="Demographic characteristics of participants in the study.", message=FALSE, warning=FALSE, out.width='100%', fig.width = 9, fig.height = 3.5}
library(patchwork)
valid.users$userAppStartTime <- as.POSIXct(valid.users$userAppStartTime,
           origin = '1970-1-1')

age.levels = sort(unique(valid.users$age))
p1 = valid.users[valid.users[,'userAppStartTime'] <= '2023-05-22',] %>% 
  group_by(age) %>% 
  summarise(count = n()) %>% 
  ggplot(mapping = aes(x = factor(age, levels = rev(age.levels)), y = count)) + 
  geom_bar(stat = 'identity') +
  geom_text(aes(x = factor(age, levels = age.levels), y = count+.5, label = age), hjust = 0) + 
  coord_flip() + 
  labs(title = 'Age', y = 'Count') +
  theme_bw() + 
  scale_y_continuous(expand = expand_scale(mult = 0, add = c(.25, 4))) + 
  theme(axis.title.y = element_blank(), axis.text.y = element_blank(), axis.ticks.y = element_blank())

p2 = valid.users[valid.users[,'userAppStartTime'] <= '2023-05-22',] %>% 
  group_by(gender) %>% 
  summarise(count = n()) %>% 
  ggplot(mapping = aes(x = gender, y = count)) + 
  geom_bar(stat = 'identity') +
  geom_text(aes(x = gender, y = count + .5, label = gender), hjust = 0) + 
  coord_flip() + 
  labs(title = 'Gender', y = 'Count') +
  theme_bw() + 
  scale_y_continuous(expand = expand_scale(mult = 0, add = c(.25, 4))) + 
  theme(axis.title.y = element_blank(), axis.text.y = element_blank(), axis.ticks.y = element_blank())

educ.levels = sort(unique(valid.users$education))[c(2,4,3,1)]
p3 = valid.users[valid.users[,'userAppStartTime'] <= '2023-05-22',] %>% 
  group_by(education) %>% 
  summarise(count = n()) %>% 
  ggplot(mapping = aes(x = factor(education, levels = educ.levels), y = count)) + 
  geom_bar(stat = 'identity') +
  coord_flip() + 
  geom_text(aes(x = education, y = count + .5, label = str_replace(education, "Graduate Degree", "Graduate\nDegree")), hjust = 0) + 
  scale_y_continuous(expand = expand_scale(mult = 0, add = c(.25, 15))) + 
  theme_bw() + 
  theme(axis.title.y = element_blank(), axis.text.y = element_blank(), axis.ticks.y = element_blank()) +
  labs(title = 'Education', y = 'Count')

@@

While the results of any sample with recruitment methodology like this are not generalizable to the public, running our initial study on a comparable population to that of the initial study serves a purpose: other replications, such as @heerCrowdsourcingGraphicalPerception2010b, used online samples of paid participants, who likely have less experience using charts and graphics and making numerical estimations than those in a statistics department (and individuals who cohabitate with them). We have every intention of running this experiment in other populations, but the convenience of online sampling is not compatible with physical objects such as our 3D printed charts, so we will have to recruit in-person participants. Thus, an initial study in a population comparable to @clevelandGraphical1984 is a reasonable first step into testing 3D charts.

@@

A Shiny applet was used for data collection, along with the provided kit of 3D printed charts.
Participants provided informed consent through the applet, and then were asked for demographic information (age, gender, education level). 
Then, participants were shown a "practice" page which allowed them to experiment with the data collection interface and practice estimating the ratio between the bars, as shown in \Cref{fig:practice}. This practice interface did not provide any feedback as to participant correctness, but was merely intended to familiarize the participants with the questions which would be asked as well as the process of estimating the ratio between the two bars.

@@

After the practice screen, participants were asked to provide the kit ID. Participants were also instructed that when indicated, they should choose a 3D chart from the kit and select the ID code of the chart (inscribed on the bottom) from the drop-down menu before completing the estimation task. 
Participants were also instructed to make quick judgements for each graph and not to measure or estimate ratios using physical objects.

@@

Each graph (or prompt, in the case of 3D printed charts) in the applet had two corresponding questions for participants to answer: first, participants were to identify the smaller bar by shape, and then, participants were to estimate the ratio of the size of the smaller bar to the size of the larger bar, as shown in \Cref{fig:experiment3dRender}. 
It should be noted that we chose to use a slider for entering ratio estimates - in part, this was intended to alleviate rounding effects which occur when participants enter numbers on their own or get numeric feedback from slider entries [@ruudUncertaintyCausesRounding2014a;@maineriSliderBarsMultiDevice2021]. 
Another benefit of this modification is that it should reduce participants' cognitive load, allowing them to focus on the ratio between bars rather than the numerical translation process.
This differs from @clevelandGraphical1984, but maintains the spirit of the study while using methods which were not available for data entry at the time.

@@

All responses ($n = `r nrow(res_all) - nrow(res)`$) that incorrectly identified the smaller bar were removed from the study before analysis; this serves as a basic attention check.

## Midmeans of Log Absolute Errors
\citeauthor{clevelandGraphical1984} used 
$$\log_2(|\text{Judged Percent} - \text{True Percent}|+1/8)$$
to measure accuracy of their participant's responses. 
In their study, log base 2 seemed appropriate due to "average relative errors changing by factors less than 10."
They also added 1/8 to prevent distortions when the errors were close to zero. 
@heerCrowdsourcingGraphicalPerception2010b followed the same analysis method, replicating many (but not all) of the results presented in the original paper.

\Cref{fig:midmeans-log-errors} shows the midmeans of the log absolute errors compared to the true ratio of the bars for each graph type and comparison type. 
The results suggest that the log absolute errors increase for greater differences between the smaller and larger bars, but are consistent across the graph types. In addition, there is a tendency

@@

res %>% 
  mutate(type = "Raw Data") %>%
  bind_rows(tmp) %>%
  # group_by(Height, graphtype, plot) %>% 
  ggplot(mapping = aes(x = Height*100, y = response,
                       color = graphtype, fill = graphtype)) +
  geom_smooth(se = T, alpha=.125,  expand = F, method = "loess") +
  geom_point(aes(shape = type, alpha = type, size = type)) +
  scale_shape_manual("Data", values = c(19, 1)) + 
  scale_size_manual("Data", values = c(2, 1)) + 
  scale_alpha_manual("Data", values = c(1, .25)) + 
  scale_fill_discrete(labels = c('Adjacent', 'Separated')) +
  scale_color_discrete(labels = c('Adjacent', 'Separated')) +
  scale_x_continuous(limits = c(0, 100)) + 
  facet_wrap(~plot, labeller = labeller(plot = plot.types)) +
  labs(title = '',
       x = 'True Proportional Difference (%)',
       y = 'Log Error',fill = 'Comparison Type',
       color = 'Comparison Type') +
  theme_bw() +
  theme(legend.position = 'bottom')
```
This is somewhat different than the results in @clevelandGraphical1984 and @heerCrowdsourcingGraphicalPerception2010b; in both cases the midmean log absolute errors increased until about 55% of the true proportional difference and then decreased.
It is possible that this difference is due to the fact that we did not require an explicit numerical estimate of the ratio but instead asked participants to indicate the ratio on a slider (which is essentially a number line). 
It is possible that this change explains the lack of a difference between chart modalities we see, but there are other potential explanations as well. 
One interesting facet of this data is that there is a consistent reduction in log error when the true proportional difference is near 50%. 
This may be because of implicit anchoring - we can typically bisect a span relatively accurately, which means that it might be expected that when the true proportion is near 50%, we would both notice that and be able to accurately transfer that information to the slider. 
This pattern is more noticable in separated (Type 3) comparisons, which also makes sense - as these comparisons are known to be less accurate, a locally  minimal error near 50% with increased error relative to adjacent (Type 1) comparisons could easily explain the exaggerated trends we see in \Cref{fig:midmeans-log-errors}.

@@

$$y_{ijklm}=\mu+S_i + R_j + G_k +T_l + \epsilon_{ijklm}$$

@@

- $S_i\sim N(0,\sigma^2_S)$ is the random effect of the $i^{th}$ subject

@@

- $G_k$ is the effect of the $k^{th}$ display type

@@

```{r, include=T}
library(papaja)
library(lme4)
